---
title: "A mini systematic review in eight steps"
subtitle: "Tutorial for lecture 8 — with templates and an optional R part"
date: "2026-11-09"
date-modified: last-modified
execute:
  eval: false
resources:
  - L08-review-templates.xlsx
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
---

::: {.callout-note}
## How to use this tutorial
This tutorial turns the [lecture](../l_08_LiteratureReviews.qmd) into eight
steps you carry out yourself. Each step points back to the lecture chapter that
explains the *why*; here we focus on the *how*.

- **Example topic:** the four-day week and other forms of working time
  reduction. It is deliberately *not* the topic of the take-home examination.
  Use it to practise, then repeat the steps on your own topic.
- **Database:** Web of Science Core Collection, accessed via the
  [ZHB](https://dbis.ur.de/HDFL/resources/3366?lang=de).
- **Templates:** download the
  [workbook with all templates](L08-review-templates.xlsx) (search log,
  screening sheet, concept matrix, limitations). It works in Excel, LibreOffice
  and Numbers.
- **Optional R part:** @sec-r shows how to analyse your Web of
  Science export with the tools from the implementation lab. The code is shown
  but not run on this page, because it needs your own export.
:::

::: {.callout-warning}
## Hit counts will differ
Web of Science is updated continuously, and the indexes available depend on the
university's subscription. Your numbers will not match anyone else's exactly —
which is why you record the **date** of every search.
:::

# Step 1: From topic to question and concept table

*Lecture chapter 2.*

A topic is not a question. Start by writing a question that tells you what is
in and what is out:

> **Example question:** *Which effects of reducing working time — in particular
> through a four-day week — on employees and firms are reported in empirical
> studies?*

Then split the question into its **concepts** and collect, for each concept,
the different ways in which authors express it. One column per concept:

| Working time reduction | Outcomes | (Empirical design) |
|:-----------------------|:---------|:-------------------|
| four-day week | productivity | trial |
| 4-day week | well-being / wellbeing | survey |
| working time reduction | job satisfaction | panel |
| reduced working hours | burnout, stress | case study |
| shorter working week | absenteeism | |
| compressed work week ⚠️ | turnover | |

: Example concept table. The third column is optional: you can capture "empirical" through the string or through your inclusion criteria — but you must document which.

::: {.callout-tip collapse="true"}
## ⚠️ Same words, different interventions
A *compressed* work week usually means the **same** number of hours in fewer
days. A four-day week in the "100-80-100" sense means **fewer** hours for the
same pay. Both are called "four-day week" in parts of the literature. Decide now
whether both are in scope — and write it into your criteria. This is the
terminology problem from lecture chapter 6 in miniature: the same word can name
different things, and different words can name the same thing.
:::

**Write down your inclusion and exclusion criteria now**, before you search.
For the example:

- **Include:** empirical studies (quantitative or qualitative) of a reduction
  in working time at firm, sector or country level that report at least one
  outcome for employees or firms.
- **Exclude:** purely conceptual or normative papers; studies of part-time work
  chosen by individuals; overtime regulation without reduction of normal hours.
- **Document type:** journal articles and reviews.
- **Language:** English (a limitation you will report in Step 8).
- **Period:** no restriction.

# Step 2: Harvest terms and build a test set

*Lecture chapter 3.*

Your concept table is only as good as your vocabulary. Three sources help:

1. **Key papers you already know.** Look at their titles, abstracts and author
   keywords. Every term they use belongs in your table.
2. **A semantic search tool** (see the
   [tool list in lecture 2](../l_02_AITools.qmd#b-grounded-search)). Ask it your
   question, then look at the *vocabulary* of the papers it suggests — not at
   its summary. This is where semantic tools are genuinely useful in a
   systematic review.
3. **Neighbouring communities.** Who else works on this? Labour economists,
   occupational health researchers, management scholars, and degrowth and
   ecological economists all study working time — with different words
   ("working time reduction", "work-sharing", "sufficiency").

At the same time, build a **test set**: three to five papers that *must* be
found by a good search. Note them in the search log. You will use them in
Step 3 to check recall.

::: {.callout-important}
## Verify what AI tools suggest
Any paper a semantic tool suggests must be found in Web of Science or opened
at the publisher before it enters your test set. A paper that you cannot find
anywhere does not exist (see the
[citation-checking exercise](../ai-citation-check/index.qmd)).
:::

# Step 3: Build and test the search string in Web of Science

*Lecture chapter 3.*

## Translate the table into a string

OR within a concept, AND between concepts. In Web of Science, open
**Advanced Search** and use field tags:

| Tag | Searches | Use for |
|:----|:---------|:--------|
| `TS=` | Topic: title, abstract, author keywords, Keywords Plus | Most searches |
| `TI=` | Title only | Very precise, low recall |
| `AB=` | Abstract only | Rarely needed |
| `AK=` | Author keywords | Checking vocabulary |

: The field tags you need most often.

A first version of the example string:

```text
TS=("four-day week*" OR "4-day week*" OR "four day week*"
    OR "working time reduction" OR "work time reduction"
    OR "reduced working hours" OR "reduction in working hours"
    OR "shorter working week*" OR "shorter workweek*")
AND
TS=(productiv* OR well-being OR wellbeing OR satisfaction
    OR burnout OR stress OR absentee* OR turnover OR performance)
```

- **Quotation marks** keep phrases together.
- `*` stands for any number of characters (`productiv*` → productivity,
  productive).
- `?` stands for exactly one character, `$` for zero or one.
- Check the Web of Science help page on *search rules* for how it handles
  hyphens and word forms; it changes occasionally.

::: {.callout-tip}
## Use the search history
Run each concept as its own search (`#1`, `#2`) and combine them in a third
search (`#1 AND #2`). You then see how many records each concept contributes —
very useful when a string returns far too many or far too few results.
:::

## Test recall with your test set

Search for each test-set paper by title. Is it in your result set? (Combine
`#3 AND TI=("exact title")`.) If a paper is missing, look at its title,
abstract and keywords: which term would have caught it? Add it, rerun, and
**log the new version**.

## Apply filters — and record them

Use *Refine results* for document type, language and years. Filters are part of
your search: write each one into the search log, together with the number of
results before and after.

# Step 4: Keep a search log

*Lecture chapter 7.*

The search log is what makes your review reproducible. Use the sheet
**Search log** in the [template workbook](L08-review-templates.xlsx). For every
version of your string, record:

| Field | Example |
|:------|:--------|
| Date | 2026-11-12 |
| Database and indexes | Web of Science Core Collection (indexes as listed under *Editions*) |
| Version | v2 |
| Full string | `TS=(...) AND TS=(...)` — copied, not paraphrased |
| Filters | Document types: Article, Review; Language: English |
| Results | number of records |
| Test set found | e.g. 4 of 5 |
| Change from previous version | added "work-sharing" after test paper 3 was missed |

: What goes into the search log.

::: {.callout-note}
A string you cannot reproduce from your log is a string you cannot report. Fill
in the log *while* you search, not afterwards.
:::

# Step 5: Export and screen

*Lecture chapter 4.*

## Export

From the result list, choose **Export → Tab delimited file**, with record
content **Full Record**. If you plan to snowball in Web of Science, include
**Cited References**. Large result sets have to be exported in batches; the
export dialogue shows the maximum per batch. Keep all files in one folder
(e.g. `data/wos/`) and do not rename the columns.

You can also export to your reference manager (RIS for Zotero) — useful for
writing, but keep the tab-delimited file as the record of your search.

## Screen titles and abstracts

Use the sheet **Screening** in the template workbook — one row per record — or
let R build it for you (@sec-r-screening). For each record:

- **decision_ta:** include / exclude / unsure (unsure stays in for round 2)
- **reason_ta:** optional at this stage

## Screen full texts

For every record that survived round 1, read the full text and decide:

- **decision_ft:** include / exclude / not retrievable
- **reason_ft:** **required** for every exclusion — use a short, fixed list of
  reasons (e.g. *not empirical*, *no working-time reduction*, *no outcome
  reported*)

::: {.callout-tip}
## Ask someone to screen a sample
Exchange 20 records with a fellow student and screen each other's sample blind.
Where you disagree, your criteria are not yet precise enough. Sharpen them, and
note the change in the search log.
:::

# Step 6: Snowball from your best papers

*Lecture chapter 3.*

Choose two or three included papers that fit your question best. In Web of
Science, open each record:

- **Backward:** *Cited References* — the papers it builds on.
- **Forward:** *Citing articles* (the "Times Cited" count) — papers that build
  on it.

Screen what you find **with the same criteria**, and count the snowballed
papers separately; they get their own box in the flow diagram. A citation-based
tool (lecture 2, "connectors") can speed this up, but note which tool and which
seed papers you used.

::: {.callout-warning}
## Snowballing stays inside the community
Papers that cite each other share a vocabulary and often a perspective.
Snowballing is good at finding what your keywords missed *within* that
community, and bad at finding the community next door.
:::

# Step 7: Fill the concept matrix and synthesise

*Lecture chapter 5.*

Before reading the included papers in depth, fix the **columns** of your
extraction sheet. Use the sheet **Concept matrix** in the template workbook.
For the example:

| Column | What you record |
|:-------|:----------------|
| Intervention | Reduced hours (e.g. 100-80-100) / compressed week / other |
| Context | Country, sector, firm size |
| Design | Trial, survey, panel, case study, … |
| Selection | How did firms or employees get into the study? |
| Outcomes | Which outcomes, measured how? |
| Findings | Direction, size, conditions |
| Caveats | What makes you trust the study more or less? |

: Example columns of the concept matrix.

Pilot the columns on three papers, adjust them once, then keep them fixed.

Then **read the matrix by column**: Where do studies agree? Where do they
disagree — and does the disagreement line up with intervention type, design or
context? That is your synthesis. Write it as a paragraph organised by these
patterns, not as one summary per paper.

::: {.callout-tip collapse="true"}
## A bias to watch for in this example
Firms that volunteer for a four-day-week trial are not a random sample of
firms. Some of the evidence also comes from organisations that campaign for
working time reduction, and appears as reports rather than journal articles.
Both are not reasons to exclude the evidence — they are things to record in
the *Selection* and *Caveats* columns and to discuss in your limitations.
:::

# Step 8: Draw the flow diagram and report

*Lecture chapters 4 and 7.*

## The flow diagram

Count what happened to every record:

| Box | From |
|:----|:-----|
| Records identified | Search log, final version |
| Records screened / excluded | Screening sheet, round 1 |
| Reports sought / not retrievable | Screening sheet, round 2 |
| Reports assessed / excluded with reasons | Screening sheet, round 2 |
| Records from other sources | Snowballing |
| Studies included | Concept matrix |

: Where the numbers for the flow diagram come from.

Draw it with the [PRISMA2020 Shiny app](https://estech.shinyapps.io/prisma_flowdiagram/),
the templates on the [PRISMA website](https://www.prisma-statement.org/), or in
R (@sec-r-prisma).

## The methods paragraph

In a term paper, the following fits into one paragraph plus an appendix:

1. Database, indexes and date of the search
2. The full final string (appendix) and all filters
3. Inclusion and exclusion criteria
4. Who screened, and how disagreements were handled
5. The numbers → flow diagram
6. How you synthesised (concept matrix, which dimensions)

## The limitations paragraph

Use the sheet **Limitations** in the template workbook. For each limitation,
name **the bias**, **its likely direction**, and **why you accepted it**.

> *Generic:* "Some relevant studies may have been missed."
>
> *Specific:* "Because the search was restricted to English-language journal
> articles in Web of Science, trial reports published by campaigning
> organisations and the German-language literature on *Arbeitszeitverkürzung*
> are not covered. Since the published trial reports tend to report positive
> outcomes, excluding them may make the review more cautious than the public
> debate; excluding the German literature likely under-represents
> sector-level studies of collective working-time agreements."

Do not forget the [AI disclosure](../ai-disclosure/index.qmd) if you used AI at
any step.

# Optional: analysing your export in R {#sec-r}

This part builds on the implementation lab sessions on
[importing data](/content/material/s_06_DataImport.qmd),
[data preparation](/content/material/s_07_DataPreparation.qmd) and
[visualisation](/content/material/s_08_Visualization.qmd). It does four
things: import and combine your exports, describe the result set (and its
biases), compare two search strings, and draw the flow diagram.

Set up an R project as usual, with your Web of Science files in `data/wos/`:

```{r}
here::i_am("R/review.R")
library(here)
library(data.table)
library(ggplot2)
```

## Import and combine the exports

Web of Science tab-delimited files use two-letter **field tags** as column
names. The ones you need most:

| Tag | Content | Tag | Content |
|:----|:--------|:----|:--------|
| `UT` | Unique Web of Science ID | `SO` | Journal (source) |
| `TI` | Title | `PY` | Publication year |
| `AU` | Authors | `DT` | Document type |
| `AB` | Abstract | `LA` | Language |
| `DI` | DOI | `WC` | Web of Science categories |
| `C1` | Author addresses | `TC` | Times cited |

: The most important Web of Science field tags.

We read every file in the folder and stack them. Reading all columns as text
(`colClasses = "character"`) avoids surprises; `quote = ""` is needed because
titles and abstracts can contain quotation marks.

```{r}
read_wos_folder <- function(folder) {
  files <- list.files(folder, pattern = "\\.txt$", full.names = TRUE)
  wos_list <- lapply(
    files, fread,
    sep = "\t", quote = "", encoding = "UTF-8",
    colClasses = "character", fill = TRUE
  )
  wos <- rbindlist(wos_list, fill = TRUE)
  unique(wos, by = "UT") # batches can overlap
}

wos <- read_wos_folder(here("data", "wos"))
nrow(wos)
```

The number of rows must equal the number of results in your search log. If it
does not, check whether a batch is missing or was exported twice.

## Describe the result set

**Publications per year.** A literature that starts suddenly in one year often
reflects a change in vocabulary (or in database coverage), not in the
phenomenon.

```{r}
wos[, PY := as.integer(PY)]

ggplot(wos[!is.na(PY), .N, by = PY], aes(x = PY, y = N)) +
  geom_col() +
  labs(x = "Publication year", y = "Records",
       title = "Records per year in the Web of Science result set") +
  theme_minimal()
```

**Journals and disciplines.** Which communities does your search reach? The
`WC` field holds the Web of Science categories, separated by semicolons.

```{r}
head(wos[, .N, by = SO][order(-N)], 15)

categories <- wos[, .(category = trimws(unlist(strsplit(WC, ";")))), by = UT]
head(categories[, .N, by = category][order(-N)], 15)
```

If almost everything comes from one or two categories (say, *Management* and
*Industrial Relations & Labor*), your string may be speaking only one
community's language.

**Languages and countries.** Two quick checks for coverage bias:

```{r}
wos[, .N, by = LA][order(-N)]

# Country of each address: the last comma-separated element; US addresses
# end in "<state> <zip> USA"
countries <- wos[, .(address = unlist(strsplit(C1, "; (?=\\[)", perl = TRUE))),
                 by = UT]
countries[, address := sub("^\\[[^]]*\\] ", "", address)]
countries[, country := trimws(sub(".*,", "", address))]
countries[grepl("USA$", country), country := "USA"]
countries <- unique(countries[country != "", .(UT, country)])
head(countries[, .N, by = country][order(-N)], 15)
```

The last table counts records with at least one author from a country. Look at
which regions are missing — and whether that matters for your question.

## Compare two search strings

Terminology bias becomes visible when you compare strings built from different
vocabularies. Export the results of two versions into two folders, e.g. a
string using only "four-day week" terms (`data/wos/string_a/`) and one that adds
"working time reduction" terms (`data/wos/string_b/`):

```{r}
string_a <- read_wos_folder(here("data", "wos", "string_a"))
string_b <- read_wos_folder(here("data", "wos", "string_b"))

only_a <- setdiff(string_a$UT, string_b$UT)
only_b <- setdiff(string_b$UT, string_a$UT)
both   <- intersect(string_a$UT, string_b$UT)
c(only_a = length(only_a), only_b = length(only_b), both = length(both))

# What does the second vocabulary add?
head(string_b[UT %in% only_b, .(PY, SO, TI)][order(-PY)], 20)
```

Read the titles that only string B finds. Are they relevant? From which
journals and disciplines do they come? This is the most direct way to see what
a single vocabulary would have cost you.

## Build the screening sheet {#sec-r-screening}

Instead of copying records by hand, write the screening sheet directly from the
export. The empty columns are filled in during screening.

```{r}
screening <- wos[, .(
  UT, AU, PY, TI, SO, DI, AB,
  decision_ta = "", reason_ta = "",
  decision_ft = "", reason_ft = "", notes = ""
)]

dir.create(here("output"), showWarnings = FALSE)
fwrite(screening, here("output", "screening_sheet.csv"), bom = TRUE)
```

::: {.callout-note collapse="true"}
## Excel shows everything in one column?
Excel with German settings expects semicolons as separators. Write the file
with `fwrite(..., sep = ";", bom = TRUE)` instead, and read it back with
`fread(..., sep = ";")`.
:::

Once screening is done, read the completed sheet back in and count:

```{r}
screened <- fread(here("output", "screening_sheet.csv"), encoding = "UTF-8",
                  colClasses = "character")

screened[, .N, by = decision_ta]
screened[decision_ta != "exclude", .N, by = decision_ft]
screened[decision_ft == "exclude", .N, by = reason_ft][order(-N)]
```

## Draw the flow diagram {#sec-r-prisma}

The [PRISMA2020](https://cran.r-project.org/package=PRISMA2020) package
[@haddaway2022prisma2020] draws the diagram from a template that ships with the
package. Install it once with `install.packages("PRISMA2020")`.

The template is a table in which the column `data` names each box and the
column `n` holds its number. Print it once to see all box names:

```{r}
library(PRISMA2020)

template <- read.csv(system.file("extdata", "PRISMA.csv", package = "PRISMA2020"))
template[, c("data", "description", "n")]
```

The small helper below sets one number and stops with an error if a box name is
misspelled — otherwise a typo would silently leave the template value in your
diagram.

```{r}
set_n <- function(tab, field, value) {
  if (!field %in% tab$data) stop("Unknown box in PRISMA template: ", field)
  tab$n[tab$data == field] <- value
  tab
}
```

Now fill in your numbers. Most of them come straight from the completed
screening sheet; only the snowballing numbers are entered by hand (the ones
below are **illustrative**). Exclusion reasons are passed as
`"reason, n; reason, n"`.

```{r}
# Database arm: computed from the screening sheet
sought       <- screened[decision_ta != "exclude", .N]
not_retrieved <- screened[decision_ft == "not retrievable", .N]
reasons      <- screened[decision_ft == "exclude", .N, by = reason_ft]
included_db  <- screened[decision_ft == "include", .N]

# Snowballing arm: entered by hand (illustrative numbers)
snowball_found    <- 9
snowball_included <- 6

counts <- list(
  database_results           = nrow(wos),
  duplicates                 = 0,
  records_screened           = nrow(screened),
  records_excluded           = screened[decision_ta == "exclude", .N],
  dbr_sought_reports         = sought,
  dbr_notretrieved_reports   = not_retrieved,
  dbr_assessed               = sought - not_retrieved,
  dbr_excluded               = paste0(reasons$reason_ft, ", ", reasons$N,
                                      collapse = "; "),
  citations_results          = snowball_found,
  other_sought_reports       = snowball_found,
  other_notretrieved_reports = 0,
  other_assessed             = snowball_found,
  other_excluded             = paste0("Not eligible, ",
                                      snowball_found - snowball_included),
  new_studies                = included_db + snowball_included,
  new_reports                = included_db + snowball_included
)

for (box in names(counts)) {
  template <- set_n(template, box, counts[[box]])
}

flow <- PRISMA_data(template)
PRISMA_flowdiagram(flow, previous = FALSE, other = TRUE, fontsize = 12)
```

`other = TRUE` draws the right-hand arm for snowballing; `previous = FALSE`
hides the arm for earlier versions of a review, which you do not have.

::: {.callout-tip}
## For the curious: bibliometrics
If you want to go further — co-citation networks, keyword maps, author
collaboration — look at the R package
[bibliometrix](https://www.bibliometrix.org/), which reads Web of Science
exports directly. It is not needed for the take-home examination.
:::

# References

::: {#refs}
:::
