---
title: "A mini systematic review in eight steps"
subtitle: "Tutorial for lecture 8 — open tools only, with templates and an optional R part"
date: "2026-11-09"
date-modified: last-modified
execute:
  eval: false
resources:
  - L08-review-templates.xlsx
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
---

::: {.callout-note}
## How to use this tutorial
This tutorial turns the [lecture](../l_08_LiteratureReviews.qmd) into eight
steps you carry out yourself. Each step points back to the lecture chapter that
explains the *why*; here we focus on the *how*.

- **Example topic:** the four-day week and other forms of working time
  reduction. It is deliberately *not* the topic of the take-home examination.
  Use it to practise, then repeat the steps on your own topic.
- **Database:** [OpenAlex](https://openalex.org) — free, open, and the same for
  everyone. Everything in the main steps works without any paid subscription.
- **Templates:** download the
  [workbook with all templates](L08-review-templates.xlsx) (search log,
  screening sheet, concept matrix, limitations). It works in Excel, LibreOffice
  and Numbers.
- **Digression:** @sec-wos shows what Web of Science adds if you have access
  through a university.
- **Optional R part:** @sec-r runs the whole search from R, which makes it fully
  reproducible. The code is shown but not run on this page, because it needs
  your own (free) API key.
:::

::: {.callout-warning}
## Hit counts will differ
OpenAlex is updated continuously. Your numbers will not match anyone else's
exactly — which is why you record the **date** and the **exact query** of every
search, and why the R part saves a copy of each result set.
:::

# Step 1: From topic to question and concept table

*Lecture chapter 2.*

A topic is not a question. Start by writing a question that tells you what is
in and what is out:

> **Example question:** *Which effects of reducing working time — in particular
> through a four-day week — on employees and firms are reported in empirical
> studies?*

Then split the question into its **concepts** and collect, for each concept,
the different ways in which authors express it. One column per concept:

| Working time reduction | Outcomes | (Empirical design) |
|:-----------------------|:---------|:-------------------|
| four-day week, 4-day week | productivity, productive | trial |
| working time reduction, work time reduction | well-being, wellbeing | survey |
| reduced working hours | job satisfaction | panel |
| shorter working week, shorter workweek | burnout, stress | case study |
| compressed work week ⚠️ | absenteeism, turnover | |

: Example concept table. The third column is optional: you can capture "empirical" through the string or through your inclusion criteria — but you must document which.

OpenAlex finds plural forms automatically (stemming), but not spelling variants
or related word forms. List them explicitly: *well-being* and *wellbeing*,
*labor* and *labour*, *productivity* and *productive*.

::: {.callout-tip collapse="true"}
## ⚠️ Same words, different interventions
A *compressed* work week usually means the **same** number of hours in fewer
days. A four-day week in the "100-80-100" sense means **fewer** hours for the
same pay. Both are called "four-day week" in parts of the literature. Decide now
whether both are in scope — and write it into your criteria. This is the
terminology problem from lecture chapter 6 in miniature: the same word can name
different things, and different words can name the same thing.
:::

**Write down your inclusion and exclusion criteria now**, before you search.
For the example:

- **Include:** empirical studies (quantitative or qualitative) of a reduction
  in working time at firm, sector or country level that report at least one
  outcome for employees or firms.
- **Exclude:** purely conceptual or normative papers; studies of part-time work
  chosen by individuals; overtime regulation without reduction of normal hours.
- **Document type:** journal articles and reviews.
- **Language:** English (a limitation you will report in Step 8).
- **Period:** no restriction.

# Step 2: Harvest terms and build a test set

*Lecture chapter 3.*

Your concept table is only as good as your vocabulary. Three sources help:

1. **Key papers you already know.** Look at their titles, abstracts and author
   keywords. Every term they use belongs in your table.
2. **A semantic search tool** (see the
   [tool list in lecture 2](../l_02_AITools.qmd#b-grounded-search)). Ask it your
   question, then look at the *vocabulary* of the papers it suggests — not at
   its summary.
3. **Neighbouring communities.** Who else works on this? Labour economists,
   occupational health researchers, management scholars, and degrowth and
   ecological economists all study working time — with different words
   ("working time reduction", "work-sharing", "sufficiency").

At the same time, build a **test set**: three to five papers that *must* be
found by a good search. Note them, with their DOIs, in the search log.

::: {.callout-important}
## Put closed-access papers into your test set
Include at least one or two papers from **closed-access journals of large
publishers** (Elsevier, Springer Nature). Many of these have no abstract in
OpenAlex (lecture chapter 3). If your string misses them, you know that the
missing abstracts are costing you recall — and you know to compensate in
Step 6.
:::

::: {.callout-important}
## Verify what AI tools suggest
Any paper a semantic tool suggests must be found in OpenAlex or opened at the
publisher before it enters your test set. A paper that you cannot find anywhere
does not exist (see the [citation-checking exercise](../ai-citation-check/index.qmd)).
:::

# Step 3: Build and test the search string in OpenAlex

*Lecture chapter 3.*

## Translate the table into a string

OR within a concept, AND between concepts:

| Element | Syntax | Note |
|:--------|:-------|:-----|
| Operators | `AND`, `OR`, `NOT` | Must be in **capitals** |
| Phrases | `"working time reduction"` | Double quotation marks |
| Grouping | `( … ) AND ( … )` | Parentheses set the order |
| Word forms | — | Plurals are found automatically (stemming); spelling variants and related forms must be listed |
| Wildcards | — | Not available in the web interface; list the variants instead |

: OpenAlex search syntax in brief. Check the [OpenAlex help pages](https://help.openalex.org) for the current rules.

A first version of the example string:

```text
("four-day week" OR "four day week" OR "4-day week" OR "four-day workweek"
 OR "working time reduction" OR "work time reduction"
 OR "reduced working hours" OR "reduction in working hours"
 OR "shorter working week" OR "shorter workweek")
AND
(productivity OR productive OR "well-being" OR wellbeing OR satisfaction
 OR burnout OR stress OR absenteeism OR turnover OR performance)
```

## Run it in the web interface

1. Go to [openalex.org](https://openalex.org) and paste the string into the
   search field.
2. Restrict the search to **title and abstract** if the interface offers this
   option. The default search also looks into the full text of those works for
   which OpenAlex has one — which makes the result depend on which papers happen
   to be open access.
3. Add filters for **year**, **type** (article, review) and **language**.
   Language tags in OpenAlex are not always correct; check a sample of what
   the filter removes.
4. Note the number of results, and **copy the page address (URL)** into your
   search log: it contains your full query and all filters.

::: {.callout-note}
## Interfaces change
Web interfaces are redesigned regularly. If a button described here has moved,
the logic stays the same: one query, explicit filters, a documented result. The
R part (@sec-r) runs exactly the same search through the API and is not affected
by interface changes.
:::

## Test recall with your test set

Search for each test-set paper by title or DOI. Is it in your result set? If a
paper is missing, look at it in OpenAlex:

- **It has an abstract** → your string lacks a term. Find the term in its
  title or abstract, add it, rerun, and **log the new version**.
- **It has no abstract** → only its title could have matched. Is there a term
  that authors in this field typically put into *titles*? Add it. Otherwise,
  note the paper — Step 6 will have to catch papers like it.

# Step 4: Keep a search log

*Lecture chapter 7.*

The search log is what makes your review reproducible. Use the sheet
**Search log** in the [template workbook](L08-review-templates.xlsx). For every
version of your string, record:

| Field | Example |
|:------|:--------|
| Date | 2026-11-12 |
| Database and route | OpenAlex, web interface (or: API via R) |
| Version | v2 |
| Full string | `( … ) AND ( … )` — copied, not paraphrased |
| Search field | Title and abstract |
| Filters | Type: article, review; language: English |
| Results | number of records |
| Query URL / script | the page URL, or the name of your R script |
| Test set found | e.g. 4 of 5 |
| Change from previous version | added "work-sharing" after test paper 3 was missed |

: What goes into the search log.

::: {.callout-note}
A string you cannot reproduce from your log is a string you cannot report. Fill
in the log *while* you search, not afterwards.
:::

# Step 5: Export, fill in missing abstracts, and screen

*Lecture chapter 4.*

## Export

Use the **export** function of the web interface and choose the spreadsheet
format (CSV). Keep this file untouched as the record of your search, and work
on a copy. Alternatively, let R fetch the records and build the screening sheet
(@sec-r-screening). For writing, you can additionally export in RIS format for
Zotero.

## Fill in missing abstracts

Before screening, sort your records by whether they have an abstract. For every
record without one:

1. **Open the publisher's page through the DOI** (`https://doi.org/` followed by
   the DOI). Abstracts are free to read there, even for articles behind a
   paywall — they are only missing from OpenAlex because publishers do not allow
   their redistribution (lecture chapter 6).
2. If the publisher page does not help, look for an **open-access version**:
   OpenAlex links to repository copies where it knows them, and many working
   papers are on SSRN, RePEc or the authors' websites.
3. Copy the abstract into the screening sheet and record **where** it came from
   (column `abstract_source`).

The R part can fill in some abstracts automatically from Crossref
(@sec-r-abstracts) — but not those of publishers that do not deposit abstracts
there, which includes Elsevier.

## Screen titles and abstracts

Use the sheet **Screening** in the template workbook — one row per record. For
each record:

- **decision_ta:** include / exclude / unsure (unsure stays in for round 2)
- **reason_ta:** optional at this stage

## Screen full texts

For every record that survived round 1, read the full text and decide:

- **decision_ft:** include / exclude / not retrievable
- **reason_ft:** **required** for every exclusion — use a short, fixed list of
  reasons (e.g. *not empirical*, *no working-time reduction*, *no outcome
  reported*)

::: {.callout-tip}
## Ask someone to screen a sample
Exchange 20 records with a fellow student and screen each other's sample blind.
Where you disagree, your criteria are not yet precise enough. Sharpen them, and
note the change in the search log.
:::

# Step 6: Snowball from your best papers — compulsory

*Lecture chapter 3.*

Snowballing is optional in many review guides. **Here it is compulsory**,
because it compensates for the missing abstracts: citation links do not depend
on abstracts, so it finds exactly those papers that a title-and-abstract search
could not see.

Choose two or three included papers that fit your question best — ideally
including one from a closed-access journal. In OpenAlex, open each record:

- **Backward:** its **references** — the papers it builds on.
- **Forward:** **cited by** — papers that build on it.

Screen what you find **with the same criteria**, and count the snowballed
papers separately; they get their own box in the flow diagram. In R,
`oa_snowball()` does both directions in one call (@sec-r-snowball).

::: {.callout-warning}
## Snowballing stays inside the community
Papers that cite each other share a vocabulary and often a perspective.
Snowballing is good at finding what your keywords missed *within* that
community, and bad at finding the community next door. That is why it
complements the string search and never replaces it.
:::

# Step 7: Fill the concept matrix and synthesise

*Lecture chapter 5.*

Before reading the included papers in depth, fix the **columns** of your
extraction sheet. Use the sheet **Concept matrix** in the template workbook.
For the example:

| Column | What you record |
|:-------|:----------------|
| Intervention | Reduced hours (e.g. 100-80-100) / compressed week / other |
| Context | Country, sector, firm size |
| Design | Trial, survey, panel, case study, … |
| Selection | How did firms or employees get into the study? |
| Outcomes | Which outcomes, measured how? |
| Findings | Direction, size, conditions |
| Caveats | What makes you trust the study more or less? |

: Example columns of the concept matrix.

Pilot the columns on three papers, adjust them once, then keep them fixed.

Then **read the matrix by column**: Where do studies agree? Where do they
disagree — and does the disagreement line up with intervention type, design or
context? That is your synthesis. Write it as a paragraph organised by these
patterns, not as one summary per paper.

::: {.callout-tip collapse="true"}
## A bias to watch for in this example
Firms that volunteer for a four-day-week trial are not a random sample of
firms. Some of the evidence also comes from organisations that campaign for
working time reduction, and appears as reports rather than journal articles.
Neither is a reason to exclude the evidence — both are things to record in the
*Selection* and *Caveats* columns and to discuss in your limitations.
:::

# Step 8: Draw the flow diagram and report

*Lecture chapters 4 and 7.*

## The flow diagram

Count what happened to every record:

| Box | From |
|:----|:-----|
| Records identified | Search log, final version |
| Records screened / excluded | Screening sheet, round 1 |
| Reports sought / not retrievable | Screening sheet, round 2 |
| Reports assessed / excluded with reasons | Screening sheet, round 2 |
| Records from other sources | Snowballing |
| Studies included | Concept matrix |

: Where the numbers for the flow diagram come from.

Draw it with the [PRISMA2020 Shiny app](https://estech.shinyapps.io/prisma_flowdiagram/),
the templates on the [PRISMA website](https://www.prisma-statement.org/), or in
R (@sec-r-prisma).

## The methods paragraph

In a term paper, the following fits into one paragraph plus an appendix:

1. Database, route (web interface or API) and date of the search
2. The full final string and all filters (appendix), plus the query URL or script
3. Inclusion and exclusion criteria
4. How missing abstracts were handled
5. Who screened, and how disagreements were handled
6. The numbers → flow diagram
7. How you synthesised (concept matrix, which dimensions)

## The limitations paragraph

Use the sheet **Limitations** in the template workbook. For each limitation,
name **the bias**, **its likely direction**, and **why you accepted it**.

> *Generic:* "Some relevant studies may have been missed."
>
> *Specific:* "The search covered titles and abstracts in OpenAlex. About a
> third of the records had no abstract, mostly closed-access articles from
> large publishers; for these, only titles could match. Backward and forward
> snowballing from four seed papers added six studies, two of them from such
> journals, but some recall loss for recent closed-access articles is likely.
> Restricting the review to English excludes the German-language literature on
> *Arbeitszeitverkürzung*, which probably under-represents sector-level
> studies of collective working-time agreements."

Do not forget the [AI disclosure](../ai-disclosure/index.qmd) if you used AI at
any step.

# Digression: what Web of Science adds {#sec-wos}

Everything above works without any subscription. If you have access to
**Web of Science** through a university, it is worth seeing what a curated,
commercial database does differently — and what that costs.

**What it adds:**

- **Precise search syntax:** field tags (`TS=` for title, abstract and
  keywords; `TI=` for titles only), truncation (`productiv*`) and wildcards
  (`organi?ation`), and proximity operators.
- **Search history:** run each concept as its own search (`#1`, `#2`) and
  combine them (`#1 AND #2`) — you see how much each concept contributes.
- **Abstracts** for nearly all indexed journal articles.
- **Curated journal selection** and consistent metadata.

**What it costs:**

- **Money and access:** an institutional subscription. Access depends on the
  contract of your university, differs between institutions — which also means
  that others may not be able to rerun your search — and ends when you leave.
- **Coverage:** fewer books, working papers, non-English and Global South
  publications than OpenAlex (lecture chapter 6).

The example string in Web of Science syntax:

```text
TS=("four-day week*" OR "4-day week*" OR "four day week*"
    OR "working time reduction" OR "work time reduction"
    OR "reduced working hours" OR "shorter working week*")
AND
TS=(productiv* OR well-being OR wellbeing OR satisfaction
    OR burnout OR stress OR absentee* OR turnover OR performance)
```

::: {.callout-tip}
## Exercise: compare the two databases
Run the equivalent string in both databases and compare the result sets by DOI
(@sec-r-wos shows how in R). Which records does only OpenAlex find, and which
only Web of Science? From which journals, languages and countries do they come?
This makes the coverage and metadata biases of lecture chapter 6 visible on
your own topic.
:::

# Optional: the whole search in R {#sec-r}

This part builds on the implementation lab sessions on
[importing data](/content/material/s_06_DataImport.qmd),
[data preparation](/content/material/s_07_DataPreparation.qmd) and
[visualisation](/content/material/s_08_Visualization.qmd). It uses the package
[openalexR](https://docs.ropensci.org/openalexR/) [@aria2024openalexr] to run
the search directly against the OpenAlex API. The big advantage: your script
*is* your search documentation — anyone can rerun it.

## Setup and API key

OpenAlex requires a free API key for programmatic access. Create an account at
[openalex.org](https://openalex.org) and copy your key from the account
settings. Then store it **outside your scripts**, so that you never share it by
accident: open your `.Renviron` file with `usethis::edit_r_environ()` and add a
line

```text
OPENALEX_KEY=your-key-here
```

Restart R. In your script:

```{r}
here::i_am("R/review.R")
library(here)
library(openalexR)
library(data.table)
library(ggplot2)

options(openalexR.apikey = Sys.getenv("OPENALEX_KEY"))
```

## Run the search

Write the string exactly as in the web interface. Avoid commas inside the
string: the API uses commas to separate filters.

```{r}
query <- '("four-day week" OR "four day week" OR "4-day week"
  OR "four-day workweek" OR "working time reduction" OR "work time reduction"
  OR "reduced working hours" OR "reduction in working hours"
  OR "shorter working week" OR "shorter workweek")
  AND (productivity OR productive OR "well-being" OR wellbeing OR satisfaction
  OR burnout OR stress OR absenteeism OR turnover OR performance)'
query <- gsub("\\s+", " ", query) # one line, single spaces

# First only count the hits: cheap, and good for comparing string versions
oa_fetch(
  entity = "works",
  title_and_abstract.search = query,
  type = c("article", "review"),
  language = "en",
  count_only = TRUE
)
```

When the count is plausible, fetch the records and **save a dated copy**.
OpenAlex changes over time; your saved copy is the result you report.

```{r}
works <- oa_fetch(
  entity = "works",
  title_and_abstract.search = query,
  type = c("article", "review"),
  language = "en"
)
works <- as.data.table(works)

dir.create(here("data"), showWarnings = FALSE)
saveRDS(works, here("data", paste0("openalex_", Sys.Date(), ".rds")))
nrow(works)
```

## Check the test set

```{r}
test_dois <- c(
  "10.xxxx/first-test-paper", # replace with the DOIs of your test set
  "10.xxxx/second-test-paper"
)
clean_doi <- function(x) tolower(sub("^https://doi.org/", "", x))

data.table(doi = test_dois, found = clean_doi(test_dois) %in% clean_doi(works$doi))
```

## Measure the abstract gap

How many of your hits have no abstract — and where are they concentrated?

```{r}
works[, no_abstract := is.na(abstract) | abstract == ""]
works[, mean(no_abstract)] # share without abstract

head(
  works[, .(records = .N, share_no_abstract = round(mean(no_abstract), 2)),
        by = host_organization_name][order(-records)],
  10
)

ggplot(works[, .(share = mean(no_abstract)), by = publication_year],
       aes(x = publication_year, y = share)) +
  geom_col() +
  labs(x = "Publication year", y = "Share of records without abstract",
       title = "Where the title-and-abstract search is effectively title-only") +
  theme_minimal()
```

These numbers belong in your limitations paragraph.

## Describe the result set

```{r}
# Records per year
ggplot(works[, .N, by = publication_year], aes(x = publication_year, y = N)) +
  geom_col() +
  labs(x = "Publication year", y = "Records") +
  theme_minimal()

# Journals and languages
head(works[, .N, by = source_display_name][order(-N)], 15)
works[, .N, by = language][order(-N)]
```

If almost everything comes from a handful of journals of one field, your
string may be speaking only one community's language.

## Compare two search strings

Terminology bias becomes visible when you compare strings built from different
vocabularies, e.g. one with only "four-day week" terms and one that adds
"working time reduction" terms:

```{r}
fetch_ids <- function(q) {
  res <- oa_fetch(entity = "works", title_and_abstract.search = q,
                  type = c("article", "review"), language = "en")
  as.data.table(res)
}
string_a <- fetch_ids('"four-day week" OR "four day week" OR "4-day week"')
string_b <- works # the full string from above

only_b <- setdiff(string_b$id, string_a$id)
length(only_b)
head(string_b[id %in% only_b, .(publication_year, source_display_name, title)], 20)
```

Read the titles that only string B finds. This is the most direct way to see
what a single vocabulary would have cost you.

## Build the screening sheet {#sec-r-screening}

```{r}
screening <- works[, .(
  id, doi, publication_year, title, source_display_name, abstract,
  abstract_source = fifelse(no_abstract, "", "OpenAlex"),
  decision_ta = "", reason_ta = "",
  decision_ft = "", reason_ft = "", notes = ""
)]
setorder(screening, abstract_source) # records without abstract first

dir.create(here("output"), showWarnings = FALSE)
fwrite(screening, here("output", "screening_sheet.csv"), bom = TRUE)
```

::: {.callout-note collapse="true"}
## Excel shows everything in one column?
Excel with German settings expects semicolons as separators. Write the file
with `fwrite(..., sep = ";", bom = TRUE)` instead, and read it back with
`fread(..., sep = ";")`.
:::

## Fill in abstracts from Crossref {#sec-r-abstracts}

Some publishers deposit abstracts with Crossref, the DOI registration agency.
The function below asks Crossref for each record without an abstract. Elsevier
does not deposit abstracts there, so for its journals you still need the
publisher page.

```{r}
crossref_abstract <- function(doi) {
  url <- paste0("https://api.crossref.org/works/", clean_doi(doi))
  res <- tryCatch(suppressWarnings(jsonlite::fromJSON(url)),
                  error = function(e) NULL) # no record or no access: NULL
  ab <- res$message$abstract
  if (is.null(ab)) return(NA_character_)
  trimws(gsub("<[^>]+>", " ", ab)) # remove the XML tags
}

missing <- screening[abstract_source == "" & !is.na(doi), id]
for (i in missing) {
  ab <- crossref_abstract(screening[id == i, doi])
  if (!is.na(ab)) {
    screening[id == i, `:=`(abstract = ab, abstract_source = "Crossref")]
  }
  Sys.sleep(0.2) # be polite to the API
}
screening[, .N, by = abstract_source]
fwrite(screening, here("output", "screening_sheet.csv"), bom = TRUE)
```

Everything still without an abstract: open the DOI link (OpenAlex stores DOIs as
links) and copy the abstract
from the publisher page by hand, setting `abstract_source` to "publisher page".

## Snowball {#sec-r-snowball}

`oa_snowball()` collects, for your seed papers, both the papers they cite and
the papers that cite them:

```{r}
seeds <- c("W0000000001", "W0000000002") # OpenAlex IDs of your seed papers

snow <- oa_snowball(identifier = seeds)
candidates <- as.data.table(snow$nodes)[oa_input == FALSE]

# Only what the string search has not already found
short_id <- function(x) sub(".*/", "", x)
new <- candidates[!short_id(id) %in% short_id(works$id)]
nrow(new)
```

Screen the new records with the same criteria and count them separately for
the flow diagram. Snowballing from highly cited papers can return thousands of
records; then restrict it, e.g. to the papers your seeds cite (backward) plus
recent citing papers.

## Draw the flow diagram {#sec-r-prisma}

Once screening is done, read the completed sheet back in:

```{r}
screened <- fread(here("output", "screening_sheet.csv"), encoding = "UTF-8",
                  colClasses = "character")

screened[, .N, by = decision_ta]
screened[decision_ta != "exclude", .N, by = decision_ft]
screened[decision_ft == "exclude", .N, by = reason_ft][order(-N)]
```

The [PRISMA2020](https://cran.r-project.org/package=PRISMA2020) package
[@haddaway2022prisma2020] draws the diagram from a template that ships with the
package. Install it once with `install.packages("PRISMA2020")`. The column
`data` names each box and the column `n` holds its number:

```{r}
library(PRISMA2020)

template <- read.csv(system.file("extdata", "PRISMA.csv", package = "PRISMA2020"))
template[, c("data", "description", "n")]
```

The small helper below sets one number and stops with an error if a box name is
misspelled — otherwise a typo would silently leave the template value in your
diagram.

```{r}
set_n <- function(tab, field, value) {
  if (!field %in% tab$data) stop("Unknown box in PRISMA template: ", field)
  tab$n[tab$data == field] <- value
  tab
}
```

Most numbers come straight from the completed screening sheet; only the
snowballing numbers are entered by hand (the ones below are **illustrative**).
Exclusion reasons are passed as `"reason, n; reason, n"`.

```{r}
# Database arm: computed from the screening sheet
sought        <- screened[decision_ta != "exclude", .N]
not_retrieved <- screened[decision_ft == "not retrievable", .N]
reasons       <- screened[decision_ft == "exclude", .N, by = reason_ft]
included_db   <- screened[decision_ft == "include", .N]

# Snowballing arm: entered by hand (illustrative numbers)
snowball_found    <- 9
snowball_included <- 6

counts <- list(
  database_results           = nrow(screened),
  duplicates                 = 0,
  records_screened           = nrow(screened),
  records_excluded           = screened[decision_ta == "exclude", .N],
  dbr_sought_reports         = sought,
  dbr_notretrieved_reports   = not_retrieved,
  dbr_assessed               = sought - not_retrieved,
  dbr_excluded               = paste0(reasons$reason_ft, ", ", reasons$N,
                                      collapse = "; "),
  citations_results          = snowball_found,
  other_sought_reports       = snowball_found,
  other_notretrieved_reports = 0,
  other_assessed             = snowball_found,
  other_excluded             = paste0("Not eligible, ",
                                      snowball_found - snowball_included),
  new_studies                = included_db + snowball_included,
  new_reports                = included_db + snowball_included
)

for (box in names(counts)) {
  template <- set_n(template, box, counts[[box]])
}

flow <- PRISMA_data(template)
PRISMA_flowdiagram(flow, previous = FALSE, other = TRUE, fontsize = 12)
```

`other = TRUE` draws the right-hand arm for snowballing; `previous = FALSE`
hides the arm for earlier versions of a review, which you do not have.

## Compare with a Web of Science export {#sec-r-wos}

For the exercise in @sec-wos: export the Web of Science results as a
**tab-delimited file** (record content *Full Record*), save it in
`data/wos/`, and compare by DOI:

```{r}
wos_files <- list.files(here("data", "wos"), pattern = "\\.txt$", full.names = TRUE)
wos <- rbindlist(lapply(
  wos_files, fread,
  sep = "\t", quote = "", encoding = "UTF-8", colClasses = "character", fill = TRUE
), fill = TRUE)
wos <- unique(wos, by = "UT")

oa_dois  <- setdiff(clean_doi(works$doi), NA)
wos_dois <- setdiff(clean_doi(wos$DI[wos$DI != ""]), NA)

c(only_openalex = length(setdiff(oa_dois, wos_dois)),
  only_wos      = length(setdiff(wos_dois, oa_dois)),
  both          = length(intersect(oa_dois, wos_dois)))

# Where do the OpenAlex-only records come from?
head(works[clean_doi(doi) %in% setdiff(oa_dois, wos_dois), .N,
           by = .(source_display_name, language)][order(-N)], 15)
```

Records without a DOI cannot be matched this way; check a few of them by title.

::: {.callout-tip}
## For the curious: bibliometrics
If you want to go further — co-citation networks, keyword maps, author
collaboration — look at the R package
[bibliometrix](https://www.bibliometrix.org/), which can read OpenAlex data
(`oa2bibliometrix()` in openalexR) as well as Web of Science exports. It is not
needed for the take-home examination.
:::

# References

::: {#refs}
:::
