---
title: "Recap exercises: import, preparation and visualization"
date: "2026-11-13"
---

The following exercises are meant to help you recap what you learned in the areas of 
*data import*, *data preparation* and *visualization*.

- [Data for exercise 2](/content/material/Recap8b/pricing_experiment.csv)
- [Template Exercise 1](/content/material/Recap8b/r_recap_exercise1.R)
- [Template Exercise 2](/content/material/Recap8b/r_recap_exercise2.R)

## Customer Satisfaction Tracking

- [Data for exercise 1](/content/material/Recap8b/customer_satisfaction.csv)

> **Background:**
You work for a company that recently implemented new customer service protocols.
Customer satisfaction data was collected at three time points: before the change (Q1),
and at two follow-up periods (Q2, Q3). Customers rated their satisfaction with
service quality and product value on a 1-5 scale.
Each row represents one customer's ratings across all three quarters.

### Import data

Download the file `customer_satisfaction.csv`and place it in your working directory.
Import the CSV file and store it in a variable called `customers_data`.

### Prepare data


The data is currently in a *wide* format - each quarter is a separate column.
Transform it to a tidy format where:
- One column indicates the quarter (`q1`, `q2`, `q3`)
- One column indicates the metric type (`service_quality`, `product_value`)
- One column contains the rating value

::: {.callout-tip collapse="true"}
#### Hint: how to separate columns
:::

### Compute grouped averages

Calculate the mean rating for each combination of quarter and metric type.
Store this in a new table called `satisfaction_summary`.

## Visualization

Create a line plot showing how average customer satisfaction changes over 
quarters. It should have the quarters on the x-axis, the mean rating on the 
y-axis, and it should show different colors for service quality and product
value.

## Bonus: compute improvements

<!--- The full solution is made available here after the session.

# Solution


For this exercise we need the following packages:

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra) # <- only for the presentation on the webpage
```

Since the csv file is in good shape we can just read it via `data.table::fread()`
or, because it is already very clean, with `readr::read_csv()`:

```{r}
# Note: adjust to you project and best use here::here()
customers_data <- read_csv(
  file = "Recap8b/customer_satisfaction.csv")
```

This is how the current data looks like:

```{r}
head(customers_data) |>
  kable()
```

To transform it into the desired long format we use `tidyr::pivot_longer()`.
Here we can make our live easier by using the selection helper `dplyr::starts_with()`
to capture all columns starting with `q`.
But a simple application only brings us that far:

```{r}
customers_data |>
  pivot_longer(
    cols = all_of(starts_with("q"))
    ) |>
  head() |>
  kable()
```

The column `name` still contains information about both the quarter and the metric.
There are a number of ways we can separate this column.

One way would be to use the argument `names_sep` of `pivot_longer()` and then
give two column names to `names_to`. This works in our case but note that it
gives a warning that some parts of the column are dropped:

```{r}
#| warning: true
customers_long <- customers_data |>
  pivot_longer(
    cols = starts_with("q"), 
    names_sep = "_", 
    names_to = c("quarter", "metric")
    ) 
head(customers_long)
```

This is because we
are telling the function to separate the column using the token `_`, but this
token appears more than once (but we only provide two column names). In our case
this is save to ignore, but when you want to avoid it, you can use the function
`dplyr::separate()`, which allows for more sophisticate separation procedures.
In our case, we can set `extra = "merge"`, which tells the function to only 
split on the first underscore and keep everything else together:

```{r}
customers_long <- customers_data |>
  pivot_longer(
    cols = starts_with("q"),
    names_to = "full_name",
    values_to = "rating"
  ) |>
  separate(
    col = full_name, 
    into = c("quarter", "metric"), 
    sep = "_", 
    extra = "merge")

head(customers_long) |>
  kable()
```

Next we need to compute average ratings for each combination of quarter and 
metric type:

```{r}
satisfaction_summary <- customers_long |>
  group_by(quarter, metric) |>
  summarise(avg_rating = mean(rating), .groups = "drop") 

kable(satisfaction_summary)
```


> Note: there are different ways of doing this. Instead of the more explicit
`group_by()` you could have used `summarize(..., .by=c())`.

The final step is then to create the visualization using `ggplot2`.
The only thing to note is that we need to be explicit regarding the grouping
and set 

```{r}
ggplot(
  data = satisfaction_summary, 
  mapping = aes(
    x = quarter, 
    y = avg_rating, 
    colour = metric, 
    group = metric)
  ) +
  geom_point() +
  geom_line() +
  theme_light()
```


::: {.callout-tip collapse="true"}
#### Why do we need `group = metric`?

When you use `geom_line()`, `ggplot2` needs to know which points to connect. 
By default, `ggplot2` automatically creates groups based on the combination of 
all discrete variables in your data.
In our case these are the two variables `quarter` and `metric`.

So `ggplot2` creates groups based on the combination of both these variables:

1. `q1 + product_value` (1 point)
2. `q1 + service_quality` (1 point)
3. `q2 + product_value` (1 point)
4. `q2 + service_quality` (1 point)
5. `q3 + product_value` (1 point)
6. `q3 + service_quality` (1 point)

Each group has only one observation, so there's nothing to connect with a line!

What we *actually* want is to have separate lines for each metric, i.e.
one line for `product_value` across all quarters and 
one line for `service_quality` across all quarters. 

So you need to tell `ggplot2`: 
"Group by metric only, not by the combination of all discrete variables, 
which would be quarter and metric!"

:::


Of course, there are many ways to make the plot pretties, but here we may leave
it at this!

To also complete the bonus challenge, we need to compute the improvement after
the intervention, i.e. from `q1` to the other two quarters. We will do this by
computer the average for `q2` and `q3`, compute the relative change to `q1` and
then rank the data according to their relative changes.

Here are all steps conducted separately: first, create a new variable to
group periods before and after the intervention:

```{r}
step1 <- satisfaction_summary |>
  mutate(
    intervention = ifelse(test = quarter=="q1", yes = "No", no = "Yes")
    )
kable(step1)
```


Now compute the respective averages:

```{r}
step2 <- step1 |>
  summarise(
    avg_rating = mean(avg_rating), .by = c("intervention", "metric")
    )
kable(step2)
```

Make it wider:

```{r}
step3 <- step2 |>
  pivot_wider(names_from = intervention, values_from = avg_rating) 
kable(step3)
```

And now compute the relative change:

```{r}
step4 <- step3 |>
  mutate(
    absolute_change = Yes - No,
    relative_change = (absolute_change / No)*100
    ) |>
  select(-No, -Yes)
kable(step4)
```

Or, all in one call:

```{r}
satisfaction_summary |>
  mutate(
    intervention = ifelse(test = quarter=="q1", yes = "No", no = "Yes")
    ) |>
  summarise(
    avg_rating = mean(avg_rating), .by = c("intervention", "metric")
  ) |>
  pivot_wider(names_from = intervention, values_from = avg_rating) |>
  mutate(
    absolute_change = Yes - No,
    relative_change = (absolute_change / No)*100
    ) |>
  select(-No, -Yes)
```
--->

